---
title: "An Exploration of Sentiment Analysis Methods to Study Student Perspectives on Artificial Intelligence"
format: html
editor: visual
---

**Professor Elodie Fourquet, Anika Gurjar**

Colgate University, Department of Computer Science

#### Preface

Prior to following this quarto document, it may be helpful to read and complete a quick refresher on R basics. I recommend the online book:

*R for Data Science (2e)* by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund (https://r4ds.hadley.nz/)

This book provides helpful information on what packages to install in R and other techniques in R. I recommend going through the first section of the book titled "Whole Game." I have also included a more detailed overview of helpful chapters from the book if you want to briefly skim though them:

[Chapter 1 - Data Visualization]{.underline}

-   Creating and customizing plots

[Chapter 3 - Data Transformation]{.underline}

-   Filtering, selecting, counting, piping(%\>%)

[Chapter 5 - Data Tidying]{.underline}

-   Tidy data, library(tidyverse)

[Chapter 7 - Data Import]{.underline}

-   Reading text files

[Chapter 10 - Exploratory Data Analysis]{.underline}

-   Summarizing data and basic plotting with ggplot2

[Chapter 14 - Strings]{.underline}

-   String manipulation with stringr - important for filtering tokens

Additionally, a majority of the information that I am sharing in this quarto document is information that I learned from **Chapter 2 - "Sentiment analysis with tidy data"** of the online book:

*Text Mining with R: A Tidy Approach* by Julia Silge and David Robinson (https://www.tidytextmining.com/sentiment)

Throughout the quarto document, when possible, I will cite the sub-section within chapter 2 of this book that I got certain information from. For example, 2.1, 2.2, etc.

A lot of the information that I am sharing in this quarto document is also information that I learned from **Chapter 6 - "From Text to Numbers"** of the book:

*Mapping Texts: Computational Text Analysis for the Social Sciences* by Dustin S. Stoltz and Marshall A. Taylor

I will also cite this book throughout this quarto document to show where I got my information from.

#### Finding the top word frequencies of a text file:

[(I learned how to do this from chapter 6 "From Text to Numbers" of *Mapping Texts: Computational Text Analysis for the Social Sciences* by Dustin S. Stoltz and Marshall A. Taylor)]{.underline}

First, we have to open the packages that we need.

```{r}
library(tidytext)  #provides function to break text into chunks called tokens (can be chunked by word, line, etc.) and to create structured format in which each row is a token

library(dplyr)  #provides functionality in data manipulation, such as filtering text

library(stringr)  #provides functionality that simplifies string manipulation, extraction, replacement, etc.

library(ggplot2) #provides functionality for data visualization such as making graphs, plots, etc.

library(tidyr)  #provides further functionality to be able to tidy datasets by creating structured formats in which each variable that you want to measure is represented by a column and each observation from the dataset is represented by a row

```

Next, we open the text file (student1_cleaned.txt) from our downloads into R. We tell R to read the lines of the text file, and give the text file the name: text_data.

```{r}
text_data <- readLines("/Users/anikagurjar/downloads/student1_cleaned.txt")
```

Next, we create a type of data frame called a tibble from text_data. In the tibble, each row shows one line from text_data. We name this tibble: text_df.

```{r}
text_df <- tibble(line = 1:length(text_data), text = text_data)
```

Next, we load stop_words, which is a built-in data set from the tidytext package. It contains a tibble of words such as "and", "the", and "of". These words do not have much meaning on their own.

```{r}
data("stop_words")
```

Next, we begin to clean our text file in order to prepare it for text analysis. The symbols "%\>%" used below is called piping, and it is used to chain together multiple operations.

```{r}
clean_text <- text_df %>%
  unnest_tokens(word, text) %>%  #tokenize(split into words)
  filter(!word %in% stop_words$word) %>%  #remove stop words
  filter(!str_detect(word, "\\d")) %>%  #remove numbers
  filter(str_length(word) > 4) #remove words less than 4 letters
```

Lastly, we find the top 5 most frequently occurring words in the cleaned text and create a data visualization to representing them. In the case that there is a tie in the frequency of words, the data visualization may display more than 5 words.

```{r}
clean_text %>%
  count(word, sort = TRUE) %>% #counts how often each word occurs and then sorts the words from most to least frequent. outputs a tibble with columns word and n.
  top_n(5) %>% #selects top 5 most frequent words
  ggplot(aes(x = reorder(word, n), y = n)) + #creates a chart, assigning x to the word reordered by frequency and y to frequency/word count (n)
  geom_col() + #draws bar plots
  coord_flip() + #flips x and y axes to make plot horizontal
  labs(title = "Top 5 Words", x = "Word", y = "Frequency") #add labels to plot
```

While the above plot gives us a good understanding of the most frequent words, it does not show the larger themes of students thoughts about AI. This is because the top words are specific to student 1's text file. So, I made a larger text file in which I combined 12 student texts. I named this file "combined_data.txt"

I then ran the same code on combined_data.txt and produced the following plot:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/combined_data.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

clean_text %>%
  count(word, sort = TRUE) %>%
  top_n(30) %>%
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 30 Words", x = "Word", y = "Frequency")

```

The above plot can give us a better understanding of the main words and themes that students are mentioning when discussing their thoughts on AI.

#### Conducting sentiment analysis of a text file:

[(Chapters used from Text Mining with R: A Tidy Approach, by Julia Silge and David Robinson: 2.1, 2.2, 2.3, and 2.4)]{.underline}

Now we can continue our analysis of our text by beginning to conduct sentiment analysis on the text file, in order to derive meaning from the text.

I will first explain the step by step process of how to complete this sentiment analysis: First, we must open the text file in R and clean it. Then, we load the bing sentiment lexicon which is included in the tidytext package. The bing sentiment lexicon is a list of words labeled as positive or negative. Next, we create a variable called "bing_words", which only contains words that appear in both the text being analyized and the bing lexicon. Then, we count the number of times that positive and negative sentiments occur and sort them separately. Lastly, we create a bar chart that shows the total number of positive words and the total number of negative words in the text.

The complete code to go along with these steps is shown below:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/student1_cleaned.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

bing <- get_sentiments("bing")  #loads bing sentiment lexicon, which is a list of words labeled as positive or negative

bing_words <- clean_text %>%   #only keeps words from cleaned text that also appear in bing lexicon
  inner_join(bing, by = "word")

bing_counts <- bing_words %>%   # counts the number of times positive and negative sentiment words appear
  count(sentiment, sort = TRUE)

ggplot(bing_counts, aes(x = sentiment, y = n, fill = sentiment)) +          # creates a bar chart showing the total number of positive vs. negative words in text, fill = sentiment gives each bar a different color
  geom_col(show.legend = TRUE) +
  labs(title = "Positive vs. Negative Words in Text (Sentiment Count Using Bing Lexicon)",
       x = "Sentiment", y = "Count")
```

The above plot shows the amount of positive vs. negative words in student 1's text. It is best to do this type of sentiment analysis using the bing lexicon with smaller text files. This is because, if you use larger text the amount of positive vs. negative words tend to even out over time which does not help us to do any sentiment analysis. Because of this, it would not be useful to conduct the above text analysis on my large text file, "combined_data.txt"

However I will conduct the sentiment analysis using the bing lexicon a few more times to other student texts.

Student 2:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/student2_cleaned.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

bing <- get_sentiments("bing")  #loads bing sentiment lexicon, which is a list of words labeled as positive or negative

bing_words <- clean_text %>%   #only keeps words from cleaned text that also appear in bing lexicon
  inner_join(bing, by = "word")

bing_counts <- bing_words %>%   # counts the number of times positive and negative sentiment words appear
  count(sentiment, sort = TRUE)

ggplot(bing_counts, aes(x = sentiment, y = n, fill = sentiment)) +          # creates a bar chart showing the total number of positive vs. negative words in text, fill = sentiment gives each bar a different color
  geom_col(show.legend = TRUE) +
  labs(title = "Positive vs. Negative Words in Text (Sentiment Count Using Bing Lexicon)",
       x = "Sentiment", y = "Count")
```

Student 10:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/student10_cleaned.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

bing <- get_sentiments("bing")  #loads bing sentiment lexicon, which is a list of words labeled as positive or negative

bing_words <- clean_text %>%   #only keeps words from cleaned text that also appear in bing lexicon
  inner_join(bing, by = "word")

bing_counts <- bing_words %>%   # counts the number of times positive and negative sentiment words appear
  count(sentiment, sort = TRUE)

ggplot(bing_counts, aes(x = sentiment, y = n, fill = sentiment)) +          # creates a bar chart showing the total number of positive vs. negative words in text, fill = sentiment gives each bar a different color
  geom_col(show.legend = TRUE) +
  labs(title = "Positive vs. Negative Words in Text (Sentiment Count Using Bing Lexicon)",
       x = "Sentiment", y = "Count")
```

Student 11:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/student11_cleaned.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

bing <- get_sentiments("bing")  #loads bing sentiment lexicon, which is a list of words labeled as positive or negative

bing_words <- clean_text %>%   #only keeps words from cleaned text that also appear in bing lexicon
  inner_join(bing, by = "word")

bing_counts <- bing_words %>%   # counts the number of times positive and negative sentiment words appear
  count(sentiment, sort = TRUE)

ggplot(bing_counts, aes(x = sentiment, y = n, fill = sentiment)) +          # creates a bar chart showing the total number of positive vs. negative words in text, fill = sentiment gives each bar a different color
  geom_col(show.legend = TRUE) +
  labs(title = "Positive vs. Negative Words in Text (Sentiment Count Using Bing Lexicon)",
       x = "Sentiment", y = "Count")
```

The different sentiment plots that we have seen above suggest that perhaps different students have different perspectives on if AI is positive or negative.

It is important to note that we should not fully trust the bing lexicon in classifying words as postive or negative. For example, the bing lexicon classifies "funny" as a negative word; as in a funny smell, or a funny feeling. It is possible that a student may have used funny as a positive word but the bing lexicon categorized it as a negative word.

If we want, we can also see which words from the text file contributed most to the postive and negative counts. We can do so using the following code:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/student1_cleaned.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

bing <- get_sentiments("bing")

bing_words <- clean_text %>%
  inner_join(bing, by = "word")

bing_counts <- bing_words %>%
  count(sentiment, word, sort = TRUE) %>%
  ungroup()

bing_counts %>%
  group_by(sentiment) %>%
  slice_max(n, n = 10) %>% 
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(x = "Contribution to sentiment",
       y = NULL)
```

The above plot shows us that the word "funny" had the highest contribution to the negative word count in student 1's text file. However I know that in student 1's text the word "funny" was used in a positive way. In order to fix this misunderstanding I can add the word "funny" to custom stop words and filter it out of the text during the cleaning process.

The following code is edited to filter out the word "funny" as a stop word:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/student1_cleaned.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

custom_stop_words <- bind_rows(tibble(word = c("funny"),  
                                      lexicon = c("custom")), 
                               stop_words)

#we add the word "funny" to the custom stop words

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% custom_stop_words$word) %>% #we filter out the custom stop words, which includes the regular stop_words data along with the word "funny"
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

bing <- get_sentiments("bing")

bing_words <- clean_text %>%
  inner_join(bing, by = "word")

bing_counts <- bing_words %>%
  count(sentiment, word, sort = TRUE) %>%
  ungroup()

bing_counts %>%
  group_by(sentiment) %>%
  slice_max(n, n = 10) %>% 
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(x = "Contribution to sentiment",
       y = NULL)
```

Now, we can see that the plot does not show the word "funny" anymore. Now that we have filtered out the word "funny", we can go back and remake the bar chart of positive vs. negative words in the text:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/student1_cleaned.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

custom_stop_words <- bind_rows(tibble(word = c("funny"),  
                                      lexicon = c("custom")), 
                               stop_words)

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% custom_stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

bing_lexicon <- get_sentiments("bing")

bing_words <- clean_text %>%
  inner_join(bing, by = "word")

bing_counts <- bing_words %>%
  count(sentiment, sort = TRUE)

ggplot(bing_counts, aes(x = sentiment, y = n, fill = sentiment)) + 
  geom_col(show.legend = TRUE) +
  labs(title = "Positive vs. Negative Words in Text (Sentiment Count Using Bing Lexicon)",
       x = "Sentiment", y = "Count")
```

This above plot that analyzes student 1's text now displays a significantly less count of negative words than it did before filtering out the word "funny."

Another way that we can conduct sentiment analysis on a text is by finding how sentiment changes over time through the text. We can do this by breaking the text into chunks, and then analyzing the positive vs. negative sentiment in each chunk to calculate the net sentiment of each chunk. Then we can create a visualization that shows the change in net sentiment over time. The code to do this is included below, with comments in the code explaining how the process is done:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/student1_cleaned.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

custom_stop_words <- bind_rows(tibble(word = c("funny"),  
                                      lexicon = c("custom")), 
                               stop_words)

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% custom_stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

bing <- get_sentiments("bing")

bing_words <- clean_text %>%
  inner_join(bing, by = "word")

bing_counts <- bing_words %>%
  count(sentiment, sort = TRUE)

sentiment_trajectory <- clean_text %>%
  inner_join(bing, by = "word") %>%   #add sentiment label to each word
  mutate(index = line %/% 2) %>%    #divides text into chunks, every two lines
  count(index, sentiment) %>%   #counts the number of positive and negative words in each chunk
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%   #makes one column for positive, one column for negative
  mutate(net_sentiment = positive - negative)   #calculates net sentiment per chunk

ggplot(sentiment_trajectory, aes(x = index, y = net_sentiment)) +
  geom_col(fill = "steelblue") +
  labs(title = "Sentiment Over Time",
       x = "Chunk Index", y = "Net Sentiment")
```

#### Data visualization as word clouds:

[(Chapters used from Text Mining with R: A Tidy Approach, by Julia Silge and David Robinson: 2.5)]{.underline}

We can create visualizations of the most frequently occurring words in a text using the wordcloud package: (In order to do this, you will have to install the wordcloud package by running the following in R: install.packages("wordcloud"))

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/combined_data.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

library(wordcloud)

clean_text %>%
  anti_join(stop_words) %>%
  count(word) %>%
  with(wordcloud(word, n, max.words = 100))
```

I used the combined data of 12 student texts to make this word visualization so that we can see the common words and themes among student perspectives. We can also make a word visualization similar to the one above that organizes the words from top to bottom, with the more positive words on the top and the more negative words on the bottom:

```{r}
library(tidytext)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)

text_data <- readLines("/Users/anikagurjar/downloads/combined_data.txt")

text_df <- tibble(line = 1:length(text_data), text = text_data)

data("stop_words")

clean_text <- text_df %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word) %>%
  filter(!str_detect(word, "\\d")) %>%
  filter(str_length(word) > 4)

library(wordcloud)
library(tibble)

word_counts <- clean_text %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0)

# Convert to matrix for comparison.cloud
word_matrix <- as.matrix(column_to_rownames(word_counts, "word"))

comparison.cloud(word_matrix,
                 colors = c("deepskyblue3", "firebrick3"),
                 max.words = 100)
```

#### Other sentiment lexicons that can be used:

[(Chapters used from Text Mining with R: A Tidy Approach, by Julia Silge and David Robinson: 2.1)]{.underline}

So far, I have only discussed the bing sentiment lexicon that contains a list of words categorized as either positive or negative.

```{r}
library(tidytext)

get_sentiments("bing")
```

However there are also other sentiment lexicons that can be used!

There is the AFINN lexicon. This lexicon assigns words with a numerical score from -5 to 5, with a negative score indicating a more negative sentiment and a positive score indicating a more positive sentiment. You can download the AFINN lexicon and then enter the following into R to view it: get_sentiments("afinn")

There is also the nrc lexicon. This lexicon categorizes words binarily (yes or no) into categories of different sentiments such as positive, negative, anger, anticipation, disgust, fear, joy, sadness, surprise, and trust. You can download the nrc lexicon and then enter the following into R to view it: get_sentiments("nrc")
